Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-20 ind_files/06_nontechnical_discussion.tex
Add a reference

Though similar effects have been recognized in several specific models of heterogamety~\citep{veller_drift-induced_2017}, life-history evolution~\citep{gillespie_natural_1974,kuosmanen_turnover_2022}, social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,constable_demographic_2016,mcleod_social_2019}, and epidemiology~\citep{kogan_two-strain_2014,parsons_pathogen_2018,day_price_2020}, this thesis is, to the best of my knowledge, the first time it has been explicitly systematically derived and formulated in full generality using a general, unifying, `model-independent' evolutionary framework. My results thus unify these previous studies under a single conceptual banner and show that their predictions are a generic outcome in finite, fluctuating populations. Most recently, a study showed (among other things) that in a very broad class of competition models of population ecology, finite populations can directionally evolve in a manner opposite to the direction of evolution predicted in the infinite population limit, essentially due to the effects of noise-induced selection~\citep{mazzolini_universality_2022}. My thesis generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a form that highlights connections with standard results in evolutionary theory such as the Price equation and Fisher's fundamental theorem (see section \ref{sec_fun_theorems}). It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}.

The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by reducing the birth rate through arresting the cell cycle, for example), thus allowing noise-induced selection for reduced turnover to operate in the system. My derivations show analytically that such deviations from neutrality in finite populations are a generic phenomenon explained by noise-induced selection and should be expected whenever there is variation in turnover rates. Selection on turnover rates also leads to insights on life-history evolution, and these insights have been extensively reported in a recent pre-print that independently arrives at my equations for type frequencies (equation \eqref{nD_stochastic_RM}) and the change of mean fitness and turnover in the population (equations \eqref{nD_stochastic_Fisher} and \eqref{nD_stochastic_Fisher_turnover} respectively) using certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration~\citep{kuosmanen_turnover_2022}.

>>>>

Though similar effects have been recognized to play a role in specific models of heterogamety~\citep{veller_drift-induced_2017,saunders_sex_2018}, life-history evolution~\citep{gillespie_natural_1974,kuosmanen_turnover_2022}, social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,constable_demographic_2016,mcleod_social_2019}, and epidemiology~\citep{kogan_two-strain_2014,parsons_pathogen_2018,day_price_2020}, these studies are usually focused on very particular systems. This thesis is, to the best of my knowledge, the first time the effect of noise-induced selection has been explicitly systematically derived and formulated in full generality using a general `model-independent' evolutionary framework such as the Price equation (or my generalization thereof). My results thus unify these previous studies under a single conceptual banner and show that their predictions are a generic outcome in finite, fluctuating populations. Most recently, a study showed (among other things) that in a very broad class of competition models of population ecology, finite populations can directionally evolve in a manner opposite to the direction of evolution predicted in the infinite population limit, essentially due to the effects of noise-induced selection~\citep{mazzolini_universality_2022}. My thesis generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a form that highlights connections with standard results in evolutionary theory such as the Price equation and Fisher's fundamental theorem (see section \ref{sec_fun_theorems}). It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}.

The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by reducing the birth rate through arresting the cell cycle, for example), thus allowing noise-induced selection for reduced turnover to operate in the system. The seemingly small systematic deviation from neutrality presented in these models can have significant evolutionary consequences over a long time (For example, see~\cite{veller_drift-induced_2017}). My derivations show analytically that such deviations from neutrality in finite populations are a generic phenomenon explained by noise-induced selection and should be expected whenever there is variation in turnover rates. Selection on turnover rates also leads to insights on life-history evolution, and these insights have been extensively reported in a recent pre-print that independently arrives at my equations for type frequencies (equation \eqref{nD_stochastic_RM}) and the change of mean fitness and turnover in the population (equations \eqref{nD_stochastic_Fisher} and \eqref{nD_stochastic_Fisher_turnover} respectively) using certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration~\citep{kuosmanen_turnover_2022}.
